---
title: "Statistical Methods for Censored Data Analysis, Demythed"
author: "James Durant"
date: 'null'
output:
  html_document: default
  pdf_document: default
  word_document: default
---


>These things cannot be explained in detail. From one thing, know ten thousand things. You must study hard.

- Miyamoto Musashi, _Go Rin No Sho_



# Introduction - why does this matter?

```{r htmltable, echo=FALSE, message=FALSE, warning=FALSE}
require(htmltab)
require(tidyverse)
require(magrittr)
require(NADA)

url <- "https://www.fda.gov/Food/FoodborneIllnessContaminants/Metals/ucm283725.htm"
juice <- htmltab(doc = url)
x <- as.numeric(as.character(juice$`Arsenic Speciation Analysis** >> Inorganic As Concentration (AsIII + AsV) (µg/kg, ppb)`))
x[x==0] <- 0.25
cen <- is.na(x)
xhalf <- x
xsqrt <- x
xzero <- x
x[cen] <- 2
xhalf[cen] <- 1
xsqrt[cen] <- 2/sqrt(2)
cen[x==0.25] <- TRUE
xhalf[x==0.25] <- 0.0
xsqrt[x==0.25] <- 0.0
xzero[cen] <- 0

dfx <- data.frame(x=x, xhalf = xhalf, xsqrt = xsqrt, xzero = xzero, cen=cen)
# 
# myx <- function(x) x
# 
# ros_apples <- ros(dfx$x, dfx$cen, forwardT="myx", reverseT="myx")

# ros_df <- as.data.frame(ros_apples)

myapples <- gather(dfx, key = "key", value = "obs", x, xhalf, xsqrt, xzero) %>%
  mutate( key = recode(key, x = "DL", xhalf = "DL/2", xsqrt = "DL/sqrt(2)", xzero = "0%*%DL")) 

sumapples <- myapples %>%
  group_by(key) %>%
  summarise(average = signif(mean(obs),3),  `standard deviation` = signif(sd(obs),3)) 


ggplot(myapples, aes(x=obs, fill = !cen)) +geom_histogram() + 
  facet_wrap(~key, labeller = label_parsed) + 
  scale_fill_manual("Detected", values = c("Grey", "Black")) +
  geom_vline(data=sumapples, aes( xintercept = average, color = "red", linetype = "dashed"), show.legend = FALSE) +
  xlab("Inorganic Arsenic Concentration [ppb]") + geom_text(data = sumapples, x = 6, y =18, aes(label = paste("bar(x)==", average)), parse = TRUE, inherit.aes = FALSE) +
  geom_text(data = sumapples, x = 6, y =12, aes(label = paste("s ==", `standard deviation`)), parse = TRUE, inherit.aes = FALSE) + 
  ggtitle("Inorganic Arsenic Concentration in 2011 FDA Survey of Apple Juice") +
  theme_bw()

```

## Learning Objectives
1. Describe the density functions for a distribution
  - Probability density function
  - Cumulative distribution function
2. Define censored data


# Density Functions


## The thing that matters - density functions!

## Probability Density Functions
Normal Probability Density Function ($\mu = true\space mean$,  $\sigma=true \space standard \space deviation$)

$$
f(x) = p(x) = \frac{1}{\sqrt{2\pi\sigma^2}}e^{-(x-\mu)^2/(2\sigma^2)}
$$

```{r pdf}
require("tidyverse")
.x  <- seq(-6, 6, length.out = 100)

dframe <- data.frame(x = .x, y = dnorm (.x, 0, 0.5), y1 = dnorm (.x, 0, 1), y2 = dnorm(.x, 0, 2))
dframe <- gather(dframe, key, value, -x)
dframe$key <- factor(dframe$key)

levels(dframe$key) <-  c("N(0,0.5)", "N(0,1)", "N(0,2)")
.plot <- ggplot(dframe, aes(x = x, y = value, linetype=key)) + 
  geom_line(size = 1) + 
  scale_y_continuous(expand = c(0.01, 0)) + 
  xlab("x") + 
  ylab("Density") + 
  labs(title = "Normal Density Functions", 
       caption = "Normal PDF for Standard Deviation 0.5, 1 and 2") + 
  scale_linetype("Standard Deviation") + theme_bw(base_size = 14, base_family = "sans")
print(.plot)


```

## Cumulative Distribution Functions
$$
F(x) = P(X \leq x) = \int_{-\infty}^x f(x) \;dx
$$
The c.d.f. of the standard normal appears as:
$$
\Phi(z) = \int_{-\infty}^z
\frac{1}{\sqrt{2}}e^{-x^2/2}\;dx
$$


## Questions

# Censoring

## Type 1 (RIGHT)

## Type 2 (LEFT)

## Type 3 (INTERVAL)

# Kaplan-Meier


## CDF and KM

## Arthmetic Mean and KM




# Central Limit Theorem


